---
title: "Authorial contributions of rOpenSci packages"
author: "Mark Padgham"
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: false
        toc_float: false
        number_sections: true
        theme: flatly
        pandoc_args: [
            "--number-offset=1"
                    ]
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  out.width = "100%",
  collapse = TRUE,
  comment = "#>",
  fig.path = "",
  echo = TRUE
)
```


A response to the [call for
help](https://github.com/ropenscilabs/annual-report-help) contributing to the
rOpenSci 2019 Annual Report, via [issue
#4](https://github.com/ropenscilabs/annual-report-help/issues/4), which aimed
to quantify "Number of authors per package". From the original description by [\@sckott](https://github.com/sckott):

> Does the number of maintainers per package change through time?

> In the interest of software sustainability, ideally each package would have
more than one maintainer, but this is relatively rare. We'd like to see the
number of maintainers per package increase over time, but number of maintainers
is hard to address without detailed knowledge of each repo. As a proxy, we
could count up all authors regardless of their role (not counting reviewers,
funders).

> Question becomes: Does the number of authors per package change (increase) through time?

Suggested approaches were based on extracting "official" authors from package
DESCRIPTION files, but numbers of authors in this context are unavoidably
cumulative, and must increase across time because authors are very generally
*not* ever removed once added. Addressing the issue through official
DESCRIPTION files would thus require comparison of these rates of increase with
some kind of neutral, expected value, which seems impracticable, so alternative
approaches are pursued here.

The analyses mostly work via several functions defined in `function-defs.R`,
loaded here, along with necessary libraries.

```{r libs, message = FALSE}
library (jsonlite)
library (dplyr)
library (magrittr)
library (ggplot2)
source ("function-defs.R")
```

The functions mostly use the github graphql API to extract the full commit
histories of all rOpenSci package repos. Not all of these are directly hosted
on `github/ropensci`, and so github organisation names also need to be
extracted.

```{r}
x <- fromJSON("https://raw.githubusercontent.com/ropensci/roregistry/gh-pages/registry.json")
pkg_names <- filter(x$packages, on_cran) %>% .$name
urls <- filter(x$packages, on_cran) %>% .$github
urls <- gsub ("https://github.com/", "", urls)
orgs <- vapply (urls, function (i) strsplit (i, "/") [[1]] [1], character (1))
repos <- vapply (urls, function (i) strsplit (i, "/") [[1]] [2], character (1))
names (orgs) <- names (repos) <- repos
```
That produces two vectors of `orgs` and `repos`, so each rOpenSci package can
be identified on github via `org/repo`. The extraction from the github graphql
API requires a client to be established with the following code:
```{r gh-cli}
token <- Sys.getenv("GITHUB_GRAPHQL_TOKEN") # or whatever
gh_cli <- ghql::GraphqlClient$new (
    url = "https://api.github.com/graphql",
    headers = httr::add_headers (Authorization = paste0 ("Bearer ", token))
)
```

## Get commit history from github


The following code extracts all of the commit histories, and takes about 20 min
or so to run:

```{r summary-fakey, eval = FALSE, echo = TRUE}
get_all_commits <- function (gh_cli)
{
    st0 <- Sys.time ()
    lapply (seq (repos), function (i) {
                   res <- process_commit_history (gh_cli, orgs [i], repos [i])
                   st <- difftime (Sys.time (), st0, units = "s") / i
                   # st is time per repo to that point
                   st <- hms::as_hms (round (st * (length (repos) - i)))
                   message (repos [i], ": ", i, " / ", length (repos),
                            "; estimated time left = ", st)
                   saveRDS (res, "results.Rds")
                   return (res)
       })
}
system.time (
             res <- get_all_commits (gh_cli)
             )
names (res) <- repos
saveRDS (res, "results.Rds")
```

## Analyse increases in numbers of contributors over time

In lieu of analysing official authors, which is untenable for reasons explained
above, the following code analyses contributions in terms both of numbers of
commits and numbers of lines of code committed. Numbers of contributors are
quantified through aggregating numbers of both commits and lines of code over
a defined time period -- fixed at 3 months throughout all of the following,
although could be easily modified -- and grouping by unique contributor.
Contributions from the primary contributor are removed from the analysis, so as
only to count contributions from additional people other than the primary
author. (It could be that the primary contributor is not the designated package
maintainer, but that situation is neither considered nor addressed here.) The
numbers are then converted to relative amounts for each time period, sorted in
decreasing order, and then converted to a linear rate of decrease per
additional unique contributor. This metric is derived for each package for each
quarter in which sufficient data are available. 

Ordered contributions should generally decrease, but *lower* rates of decrease
are taken here to imply greater numbers of additional contributors. While this
is not necessarily directly so, it appears to us to provide the most direct
insight into relative contributions of multiple authors, as depicted in the
figure below.

```{r rel-contr-ex, echo = FALSE}
x1 <- 1:3
y1 <- c (3, 2, 1) / 6
y2 <- c (9, 5, 1) / 15
dat1 <- data.frame (x = c (x1, x1),
                    y = c (y1, y2),
                    type = c (rep ("a", length (x1)), rep ("b", length (x1))),
                    plot = "(A)",
                    stringsAsFactors = FALSE)
x2 <- 1:4
y3 <- c (7, 5, 3, 1) / 16
dat2 <- data.frame (x = c (x1, x2),
                    y = c (y2, y3),
                    type = c (rep ("b", length (x1)), rep ("c", length (x2))),
                    plot = "(B)",
                    stringsAsFactors = FALSE)

dat <- rbind (dat1, dat2)
```
```{r rel-contr-ex-plot, echo = FALSE}
ggplot (dat, aes (x, y, colour = type, fill = type)) +
    geom_col (position = "dodge") +
    geom_line (lwd = 3, linetype = "dashed") +
    facet_grid (.~plot) +
    theme (legend.position = c (0.9, 0.9))
```

While admittedly not particularly easy to interpret (sorry about that!), the
figure is meant to demonstrate the abiding hypothesis of this analysis. Panel
(A) shows two different relative rates of contribution, yet with both arising
from the three distinct contributors. Panel (B) depicts the main hypothesis
here, which is that reductions in the rate of decrease in relative
contributions will generally reflect more contributors -- the blue line ("c")
has a lower absolute slope because it has more contributors than the green line
("b"). It may nevertheless be that relatively lower slopes simply reflect more
relatively equal contributions -- illustrated by the red line ("a") in Panel
(A) having a lower slope than the green line ("b"), even though both reflect
three contributors. *The following analysis assume that any arrangement of
contributions by which successively greater numbers contribute relatively more
reflect greater numbers of authors contributing to a package,* where through
actual increases in number of authors, or relatively greater contributions by
those making more minor contributions, or both.

With that hopefully sufficiently clarified, let us proceed ...


# Numbers of distinct contributors over time

The following code implements one approach to identifying whether numbers of
contributors increases over time, through quantifying for each time period the
number of (commits, lines) per author, ranking them, and estimating regression
slopes for that period of time. Increases in actual numbers of contributors
must then manifest through regression slopes becoming *less negative* over time.

```{r authors-per-time-interval}
library (magrittr)
library (dplyr)
library (ggplot2)
dat <- readRDS ("results.Rds")
nr <- vapply (dat, function (i)
              ifelse (is.null (i), 0L, nrow (i)), integer (1))
dat <- dat [which (nr > 0)]
stats <- lapply (dat, function (i) {
        # convert dates to quarterly fractions for grouping below
        y <- as.integer (lubridate::year (i$date))
        qtr <- ceiling (lubridate::month (i$date) / 3)
        i$date <- y + (qtr - 1) / 4

        # filter primary contributor
        primary <- names (sort (table (i$name), decreasing = TRUE)) [1]
        i <- filter (i, name != primary)

        if (nrow (i) < 2)
            return (NULL)

        # decrease in relative contributions per author for that time period
        # first for number of commits:
        s1 <- group_by (i, date, name) %>%
            summarise (n = length (name)) %>%
            group_by (date) %>%
            mutate (len = length (n)) %>%
            filter (len > 1) %>%
            group_by (date) %>%
            mutate (n = sort (n / sum (n), decreasing = TRUE))
        if (nrow (s1) > 0)
        {
            suppressWarnings ({ # essentially perfect lm fit
                s1 <- group_by (s1, date) %>%
                summarise (slope = summary (lm (n ~ seq (n)))$coefficients [2]) %>%
                filter (is.finite (slope))
            })
        }
        if (nrow (s1) > 0)
            s1$var <- "commits"
        else
            s1 <- NULL

        # then for lines of code:
        s2 <- group_by (i, date, name) %>%
            summarise (n = sum (additions)) %>%
            group_by (date) %>%
            mutate (len = length (n)) %>%
            filter (len > 1) %>%
            group_by (date) %>%
            mutate (n = sort (n / sum (n), decreasing = TRUE))
        if (nrow (s2) > 0)
        {
            suppressWarnings ({ # essentially perfect lm fit
                s2 <- group_by (s2, date) %>%
                summarise (slope = summary (lm (n ~ seq (n)))$coefficients [2]) %>%
                filter (is.finite (slope))
            })
        }
        if (nrow (s2) > 0)
            s2$var <- "lines"
        else
            s2 <- NULL

        bind_rows (s1, s2)
    })
stats <- stats [which (!sapply (stats, is.null))]
stats <- data.frame (do.call (rbind, stats))

ggplot (stats, aes (date, slope)) +
    geom_point (colour = "lawngreen") +
    geom_smooth (method = "lm") +
    facet_grid (.~var, scales = "free")
```

Over the times analysed for the rOpenSci repos (which obviously in some cases
extend well prior to the actual beginnings of the organization), repos have on
average seen a progressive increase in the relative proportions of minor
contributions -- that is, either numbers of non-primary contributors have
increased, or the relative contributions of non-primary contributors have
become more equal, or both.

This is the closest I can get to demonstrating that the "number of maintainers
per package change[s] through time". Are these changes significant?

```{r summary-of-authors-per-time}
summary (lm (stats$slope [stats$var == "commits"] ~
             stats$date [stats$var == "commits"]))
summary (lm (stats$slope [stats$var == "lines"] ~
             stats$date [stats$var == "lines"]))
```

Yep, relative numbers of commits from successive non-primary contributors
manifest a significant increase over time. Equivalent changes in lines of code
are not significant. The left-hand panel suggests that one reason for these
increases is an increase in the number of quarters in which contributions from
non-primary contributors were all equal, yielding slopes of relative change
equal to zero. This itself suggests either very few contributors, likely only
two, or very small contributions of one or two commits, or both. These may be
readily excluded in the above code by replacing the lines
```{r, eval = FALSE}
filter (len > 1)
```
with 
```{r, eval = FALSE}
filter (len > 2)
```
... doing so, however, entirely removes the significance of any result. I would
nevertheless suggests that it is appropriate to leave these values, as they
still represent valid non-primary contributions no matter how minor.

